The Nuclear Norm and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-08-20 conf_article/neon.tex
Added L-smooth part

%------------------------------------------------------------------------------------
\newcommand{\Rmn}{\R^{m\times n}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\cS}{\mathcal{S}}
\newcommand{\cB}{\mathcal{B}}
\newcommand{\Ed}[2]{\mathbb{E}_{#1}\left[#2\right]}
\usepackage{mathtools}
\DeclarePairedDelimiter{\sqne}{\|}{\|_2^2}
%\DeclarePairedDelimiter{\norme}{\|}{\|_2}
\DeclarePairedDelimiter{\normf}{\|}{\|_\mathrm{F}}
\DeclarePairedDelimiter{\normfstar}{\|}{\|_\mathrm{F*}}
\DeclarePairedDelimiter{\normftwo}{\|}{\|_\mathrm{F2}}
\DeclarePairedDelimiter{\sqns}{\|}{\|_{\mathrm{op}}^2}
\DeclarePairedDelimiter{\norms}{\|}{\|_{\mathrm{op}}}
\DeclarePairedDelimiter{\sqnn}{\|}{\|_{\mathrm{nuc}}^2}
\DeclarePairedDelimiter{\sqnf}{\|}{\|_{\mathrm{F}}^2}
\DeclarePairedDelimiter{\normn}{\|}{\|_{\mathrm{nuc}}}
\def\<#1,#2>{\langle #1,#2\rangle}
\DeclarePairedDelimiter{\dotprod}{\langle}{\rangle}

@@

\section{Introduction}
 Minimizing loss functions in unprecedentedly high-dimensional spaces has recently become an integral and crucial part in training large language models. Hence, new scalable, time- and memory-efficient algorithms have been demanded. Besides well-known Adam and AdamW (citations), recently proposed Muon has shown promising results on different benchmarks (CIFAR, NanoGPT, citations). Its key difference from earlier algorthims is that it has been constructed specifically for optimizing functions of weight matrices. Such functions are common in modern machine learning (citations), so it does not significantly restrict Muon's applicability to modern problems.

 That is what can be said from a practical point of view. From the perspective of theory, Muon's main innovation was an intentional usage of matrix norms, i.e. the spectral norm, to derive the algorithm's update (cite Deriving Muon). Several other attempts have been since made to construct new algorithms, mainly generalizations of Muon's paradigm (cite Scion and Gluon).

 Based upon recent theoretical advances that explain some theory behind Muon, Scion and Gluon (cite), we explore application of other matrix norms to optimization of functions of matrices. As it has been done with Muon, we stipulate that our algorithms' updates be fast to compute.
\iffalse
\begin{itemize}
    \item Hype of Muon and others:
    \begin{itemize}
        \item Importance of the functions of a matrix weight. Special structure.
        \item Write the equation
        \item Memory efficiency and hyperparameter transferability
        \item Proven efficient computing (cite benchmarks and Dion and Cosmos)
    \end{itemize}
    \item Comparison:
    \begin{itemize}
        \item Shampoo (very compute intensive for large layers, we need asymptotics!), second order
        \item LMO has produced several effective algorithms: Muon, Dion, Scion, Gluon (citations)
        \item generalization is better for Muon and others (citations!). Possibly nuSAM here (definitely)
        \item Scion and Gluon for multilayer
        \item Gluon for (L0, L1)-smooth problems, which is more realistic (low L0, faster convergence - no impractically small stepsizes)
    \end{itemize}
    \item Analyis:
    \begin{itemize}
        \item Several generalizations have been already presented by the community: Old Optimizer and New Norm (idea and evidence that it explains famous algorithms); Pethick (LMO and different layers); Kovalev (as trust region and arbitrary norm); Gluon (generalized smoothness model).
        \item However, to the best of our knowledge, no attempt has been made to try other norms. Here we come in.
        \item No attempt to mitigate high computation cost of matrix computation has been made. Here again we come in.
        \item It has been shown that $O(1/esp^4)$ bounds for Muon, Scion, Gluon and even Adam for L-smooth functions (citations). Some bounds depend on the dimensionality of space (cite Kovalev), which for spectral norm gives much more iterations (cite Kovalev)
        \item This motivates us to try new norms
    \end{itemize}
\end{itemize}
\fi

@@

{\bf Non-Euclidean norm setting and Lipschitz continuous gradient.}
    We assume that matrix space $\Rmn$ is equipped with a norm $\norm{\cdot}\colon \Rmn \to \R_+$, which possibly does not coincide with the Frobenius, norm $\normf{\cdot}$. In addition, we assume that the gradient $\nabla f(\cdot)$ is Lipschitz continuous with respect to the norm $\norm{\cdot}$, that is, the following inequality holds:
    \begin{aequation}\label{eq:L}
    \norm{\nabla f(\mX) - \nabla f(\mX')}_* \leq L \norm{\mX-\mX'}
    \quad\text{for all}\;
    \mX, \mX' \in \Rmn,
    \end{aequation}
    where $L > 0$ is the gradient Lipschitz constant, and $\norm{\cdot}^\dagger\colon \Rmn \to \R_+$ is the dual norm associated with $\norm{\cdot}$, i.e., $\norm{\mX}^\dagger = \sup_{\norm{\mX'}\leq 1} \<\mX,\mX'>$ for all $\mX \in \Rmn$. The assumption of gradient Lipschitz continuity is also widespread in the analysis of first-order optimization methods \citep{ghadimi2013stochastic,gower2019sgd,cutkosky2020momentum,horvath2023stochastic,gorbunov2020linearly}. It is important to highlight that while \Cref{eq:L} uses the dual norm $\norm{\cdot}^\dagger$ to measure the difference between the gradients, the variance in \Cref{eq:variance} is measured with respect to the Frobenius norm $\sqnf{\cdot}$, which is necessary to properly utilize the unbiasedness property of the stochastic gradient estimator $g(\cdot;\xi)$. Therefore, we need to provide a connection between these norms using the following inequality:
    \begin{aequation}\label{eq:norm}
    \norm{\mX}^\dagger \leq \rho\cdot\normf{\mX}
    \quad\text{for all}\;
    \mX \in \Rmn,
    \end{aequation}
    where $\rho > 0$ is a positive constant. Note that such a constant always exists due to the norm equivalence theorem, which always holds in the finite-dimensional space $\Rmn$. We recount $\rho$ for different norms $\norm{\cdot}$ in the appendix.

@@

        Lemma. When $\norm{\cdot} = \normf{\cdot}$, \cref{eq:our_update} turns into:

@@

        It is an interesting observation, because in other works (\cite{pethick2025training}), $\normf{\cdot}$ was used to recover SGD. The difference is in how one states the problem.

    {\bf $\norms{\mM^{k}}$ and Muon}

        Lemma. When $\norm{\cdot} = \norms{\cdot}$, \cref{eq:our_update} turns into:

        \begin{equation}\label{eq:muon_update}
            \mX^{k+1} = \mX^{k} - \eta U V^\top
        \end{equation}

    {\bf $\normn{\mM^{k}}$ and Neon}

        Lemma. When $\norm{\cdot} = \normn{\cdot}$, \cref{eq:our_update} turns into:

        \begin{equation}\label{eq:neon_update}
            \mX^{k+1} = \mX^{k} - \eta u_1 v_1^\top
        \end{equation}

        We name the derived algorithm \emph{Neon}. In the section Matrix side of updates, we will discuss how to compute an update efficiently.

@@

        Lemma. When $\norm{\cdot} = \normfstar{\cdot}^\dagger$, \cref{eq:our_update} turns into:

        \begin{equation}\label{eq:nsgd_muon_update}
            \mX^{k+1} = \mX^{k} - \eta (\alpha U V^\top + (1-\alpha)\frac{\mM^k}{\normf{\mM^k}})
        \end{equation}

@@

        Lemma. When $\norm{\cdot} = \normftwo{\cdot}^\dagger$, \cref{eq:our_update} turns into:

        \begin{equation}\label{eq:nsgd_neon_update}
            \mX^{k+1} = \mX^{k} - \eta (\alpha u_1 v_1^\top + (1-\alpha)\frac{\mM^k}{\normf{\mM^k}})
        \end{equation}

@@

\section{"Smooth" theory via Trust Region bounds}
\begin{itemize}
    \item Probably statement of the trust region problem
    \item Assumption on unbiasedness and limited variance of stoch. grad.
    \item Assumption on L-smoothness as in the Kovalev's article.
    \item Assumption on rho. To be more precise, the statement of lemmas on rho for all 4 norms.
    \item We utilize the Kovalev's corollary to derive convergence
    \item Bonus: convert all the norms into a common sense (L2?) norm to get the true dependency on dimensionality
\end{itemize}

@@

\subsection{}
~ Note: nuSAM and their update. They had no Lanczos
Technical details on experiments

@@

\bibliography{icomp2024_conference}
\bibliographystyle{icomp2024_conference}

@@

>>>>

%------------------------------------------------------------------------------------
% \epsilon is predefined; override to use \varepsilon
\renewcommand{\epsilon}{\varepsilon}
\newcommand{\Rmn}{\R^{m\times n}}
\newcommand{\cB}{\mathcal{B}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\cO}{\mathcal{O}}
\newcommand{\cS}{\mathcal{S}}
\newcommand{\Ed}[2]{\mathbb{E}_{#1}\left[#2\right]}
\usepackage{mathtools}
\DeclarePairedDelimiter{\sqne}{\|}{\|_2^2}
%\DeclarePairedDelimiter{\norme}{\|}{\|_2}
\DeclarePairedDelimiter{\normf}{\|}{\|_\mathrm{F}}
\DeclarePairedDelimiter{\normfstar}{\|}{\|_\mathrm{F*}}
\DeclarePairedDelimiter{\normftwo}{\|}{\|_\mathrm{F2}}
\DeclarePairedDelimiter{\sqns}{\|}{\|_{\mathrm{op}}^2}
\DeclarePairedDelimiter{\norms}{\|}{\|_{\mathrm{op}}}
\DeclarePairedDelimiter{\sqnn}{\|}{\|_{\mathrm{nuc}}^2}
\DeclarePairedDelimiter{\sqnf}{\|}{\|_{\mathrm{F}}^2}
\DeclarePairedDelimiter{\normn}{\|}{\|_{\mathrm{nuc}}}
\def\<#1,#2>{\langle #1,#2\rangle}
\DeclarePairedDelimiter{\dotprod}{\langle}{\rangle}

@@

\section{Introduction}
 Minimizing loss functions in unprecedentedly high-dimensional spaces has recently become an integral and crucial part in training large language models. Hence, new scalable, time- and memory-efficient algorithms have been demanded. Besides well-known Adam and AdamW \cite{kingma2014adam}, \cite{loshchilov2017decoupled}, recently proposed Muon has shown promising results on different benchmarks (CIFAR, NanoGPT, citations). Its key difference from earlier algorthims is that it has been constructed specifically for optimizing functions of weight matrices. Such functions are common in modern machine learning (citations), so it does not significantly restrict Muon's applicability to modern problems.

 That is what can be said from a practical point of view. From the perspective of theory, Muon's main innovation was an intentional usage of matrix norms, i.e. the spectral norm, to derive the algorithm's update (cite Deriving Muon). Several other attempts have been since made to construct new algorithms, mainly generalizations of Muon's paradigm (\cite{pethick2025training} and \cite{riabinin2025gluon}).

 Based upon recent theoretical advances that explain some theory behind Muon, Scion and Gluon (\cite{bernstein2025deriving,kovalev2025understanding,pethick2025training,riabinin2025gluon}), we explore application of other matrix norms to optimization of functions of matrices. As it has been done with Muon, we stipulate that our algorithms' updates be fast to compute.
\iffalse
\begin{itemize}
    \item Hype of Muon and others:
    \begin{itemize}
        \item Importance of the functions of a matrix weight. Special structure.
        \item Write the equation
        \item Memory efficiency and hyperparameter transferability
        \item Proven efficient computing (cite benchmarks and Dion and Cosmos)
    \end{itemize}
    \item Comparison:
    \begin{itemize}
        \item Shampoo (very compute intensive for large layers, we need asymptotics!), second order
        \item LMO has produced several effective algorithms: Muon, Dion, Scion, Gluon (citations)
        \item generalization is better for Muon and others (citations!). Possibly nuSAM here (definitely)
        \item Scion and Gluon for multilayer
        \item Gluon for (L0, L1)-smooth problems, which is more realistic (low L0, faster convergence - no impractically small stepsizes)
    \end{itemize}
    \item Analyis:
    \begin{itemize}
        \item Several generalizations have been already presented by the community: Old Optimizer and New Norm (idea and evidence that it explains famous algorithms); Pethick (LMO and different layers); Kovalev (as trust region and arbitrary norm); Gluon (generalized smoothness model).
        \item However, to the best of our knowledge, no attempt has been made to try other norms. Here we come in.
        \item No attempt to mitigate high computation cost of matrix computation has been made. Here again we come in.
        \item It has been shown that $O(1/esp^4)$ bounds for Muon, Scion, Gluon and even Adam for L-smooth functions (citations). Some bounds depend on the dimensionality of space (cite Kovalev), which for spectral norm gives much more iterations (cite Kovalev)
        \item This motivates us to try new norms
    \end{itemize}
\end{itemize}
\fi

@@

{\bf Non-Euclidean norm setting and Lipschitz continuous gradient.}
    We assume that matrix space $\Rmn$ is equipped with a norm $\norm{\cdot}\colon \Rmn \to \R_+$, which possibly does not coincide with the Frobenius, norm $\normf{\cdot}$. In addition, we assume that the gradient $\nabla f(\cdot)$ is Lipschitz continuous with respect to the norm $\norm{\cdot}$, that is, the following inequality holds:
    \begin{aequation}\label{eq:L}
    \norm{\nabla f(\mX) - \nabla f(\mX')}^\dagger \leq L \norm{\mX-\mX'}
    \quad\text{for all}\;
    \mX, \mX' \in \Rmn,
    \end{aequation}
    where $L > 0$ is the gradient Lipschitz constant, and $\norm{\cdot}^\dagger\colon \Rmn \to \R_+$ is the dual norm associated with $\norm{\cdot}$, i.e., $\norm{\mX}^\dagger = \sup_{\norm{\mX'}\leq 1} \<\mX,\mX'>$ for all $\mX \in \Rmn$. The assumption of gradient Lipschitz continuity is also widespread in the analysis of first-order optimization methods \citep{ghadimi2013stochastic,gower2019sgd,cutkosky2020momentum,horvath2023stochastic,gorbunov2020linearly}. It is important to highlight that while \Cref{eq:L} uses the dual norm $\norm{\cdot}^\dagger$ to measure the difference between the gradients, the variance in \Cref{eq:variance} is measured with respect to the Frobenius norm $\sqnf{\cdot}$, which is necessary to properly utilize the unbiasedness property of the stochastic gradient estimator $g(\cdot;\xi)$. Therefore, we need to provide a connection between these norms using the following inequality:
    \begin{aequation}\label{eq:norm}
    \norm{\mX}^\dagger \leq \rho\cdot\normf{\mX}
    \quad\text{for all}\;
    \mX \in \Rmn,
    \end{aequation}
    where $\rho > 0$ is a positive constant. Note that such a constant always exists due to the norm equivalence theorem, which always holds in the finite-dimensional space $\Rmn$. We recount $\rho$ for different norms $\norm{\cdot}$ in the appendix.

@@

        \begin{lemma}\label{lemma:nsgd_update}
            When $\norm{\cdot} = \normf{\cdot}$, \cref{eq:our_update} turns into:

@@

        \end{lemma}
        It is an interesting observation, because in other works (\cite{pethick2025training}), $\normf{\cdot}$ was used to recover SGD. The difference is in how one states the problem.

    {\bf $\norms{\mM^{k}}$ and Muon}
    \begin{lemma}\label{lemma:muon_update}
        When $\norm{\cdot} = \norms{\cdot}$, \cref{eq:our_update} turns into:

        \begin{equation}\label{eq:muon_update}
            \mX^{k+1} = \mX^{k} - \eta U V^\top
        \end{equation}
    \end{lemma}

    {\bf $\normn{\mM^{k}}$ and Neon}
    \begin{lemma}\label{lemma:neon_update}
        When $\norm{\cdot} = \normn{\cdot}$, \cref{eq:our_update} turns into:

        \begin{equation}\label{eq:neon_update}
            \mX^{k+1} = \mX^{k} - \eta u_1 v_1^\top
        \end{equation}
    \end{lemma}
        We name the derived algorithm \emph{Neon}. In the section Matrix side of updates, we will discuss how to compute an update efficiently.

@@

    \begin{lemma}\label{lemma:nsgd_muon_update}
        When $\norm{\cdot} = \normfstar{\cdot}^\dagger$, \cref{eq:our_update} turns into:

        \begin{equation}\label{eq:nsgd_muon_update}
            \mX^{k+1} = \mX^{k} - \eta (\alpha U V^\top + (1-\alpha)\frac{\mM^k}{\normf{\mM^k}})
        \end{equation}
    \end{lemma}

@@

    \begin{lemma}\label{lemma:nsgd_neon_update}    
        When $\norm{\cdot} = \normftwo{\cdot}^\dagger$, \cref{eq:our_update} turns into:

        \begin{equation}\label{eq:nsgd_neon_update}
            \mX^{k+1} = \mX^{k} - \eta (\alpha u_1 v_1^\top + (1-\alpha)\frac{\mM^k}{\normf{\mM^k}})
        \end{equation}
    \end{lemma}

@@

\section{Trust Region Bounds for L-smooth functions}
    First, we analyze the problem in the unstochastic case. From Corollary 1 of \cite{kovalev2025understanding}, we directly get the following result that matches lower bounds, as was noted in \cite{kovalev2025understanding}.

    \begin{lemma}\label{lemma:no_noise_tr}
    To reach the precision $\min_{k=1\ldots K} \norm{\nabla f(\mX_k)}^\dagger \leq \epsilon$ by the iterations~\eqref{eq:our_update} under the conditions of \Cref{eq:L}, it is sufficient to choose the stepsize $\eta$ and the number of iterations $K$ as follows:
    \begin{equation}
      \eta = \cO\left(\frac{\epsilon}{L}\right),\qquad K = \cO\left(\frac{L\Delta_0}{\epsilon^2}\right).
      % \eta = \frac{\epsilon}{3L},\qquad K = \ceil*{\frac{6L\Delta_0}{\epsilon^2}}.
    \end{equation}
    \end{lemma}

    In the stochastic case, from Corollary of 2 of \cite{kovalev2025understanding}, we directly get the following result that matches lower bounds, as was noted in \cite{kovalev2025understanding}:

    \begin{lemma}\label{lemma:stoch_tr}
    To reach the precision $\E{\min_{k=1\ldots K} \norm{\nabla f(\mX_k)}^\dagger} \leq \epsilon$ by ~\eqref{eq:our_update} under the assumptions \Cref{eq:variance}, \Cref{eq:L}, \Cref{eq:norm}, it is sufficient to choose the parameters as follows:
    \begin{align}
        \eta &= \cO\left(\min\left\{\frac{\epsilon}{L}, \frac{\epsilon^3}{\rho^2\sigma^2L}\right\}\right),
        \qquad
        \alpha = \cO\left(\min\left\{1, \frac{\epsilon^2}{\rho^2\sigma^2}\right\}\right),
        \\
        \label{eq:str_K_nonconvex}
        K &= \cO\left(\max\left\{
            \frac{\rho\sigma}{\epsilon},
            \frac{\rho^3\sigma^3}{\epsilon^3},
            \frac{L\Delta_0}{\epsilon^2},
            \frac{L\Delta_0\rho^2\sigma^2}{\epsilon^4}
        \right\}\right).
    \end{align}
    \end{lemma}

    As the norms $\normf{\cdot}$, $\normn{\cdot}$, $\normf{\cdot}$ are almost proportional to each other when $m, n \rightarrow \infty$ (with high probability for random matrices), the expected convergence guarantees in terms of $\normf{\cdot}$ are the same (it can be easily shown by noting that $\norm{\mX}\sim\alpha\normf{\mX}$, $\norm{\nabla f(\mX)}^\dagger \sim \frac{1}{\alpha} \normf{\nabla f(\mX)}$, and expressing $L$-constant via $L_F$-constant for the Frobenius norm).

@@

\subsection{}
~ Note: nuSAM and their update. They had no Lanczos
Technical details on experiments
\iffalse

@@

@@

\fi
\bibliography{icomp2024_conference}
\bibliographystyle{icomp2024_conference}
